---
title: "Orton_29Sept_Phase1"
output: html_document
date: "2023-09-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Project 2
#5 R objects
##Atomic Vectors- a simple vector of data. is.vector() is how you as R if this object is a vector. Create an integer vector by including a capital L with the input. Put quotes around some input to make the vector a character.
```{r}
int <- 1L
text <- "ace"
```
##Doubles atomic vector-the default type for vectors. Can find the type of an object with typeof(). Doubles are accurate to 16 significant digits, which can resulting in some less precise rounding, aka floating-point arithmetic/errors and can cause different results. Still ok to generally use doubles.
##Integers- numbers that can be written without a decimal compnonent. Make this by following the input with L
```{r}
int <- c(-1L, 2L, 4L)
int
typeof(int) 
```
##Characters- stores text. Individual elements of a character vector are called strings; strings can be numbers, letters, or symbols
```{r}
text <- c("Hello",  "World")
text
##  "Hello"  "World"

typeof(text)
## "character"

typeof("Hello")
## "character"
```
##Logicals- logical vectors store TRUEs and FALSEs
```{r}
3>4

logic <- c(TRUE, FALSE, TRUE)
logic
##   TRUE FALSE  TRUE

typeof(logic)
## "logical"

typeof(F)
## "logical"
```
##Complex and Raw- complex vectors store complex numbers, like imaginary numbers. Raw vecors store raw bytes of data. Make an empty raw vector of length n with raw(n).
```{r}
comp <- c(1 + 1i, 1 + 2i, 1 + 3i)
comp
## 1+1i 1+2i 1+3i

typeof(comp)
## "complex"

raw(3)
## 00 00 00

typeof(raw(3))
## "raw"
```
##Vector of Cards
##Create a vector of cards. The most appropriate would be character vectors. 
```{r}
hand <- c("ace", "king", "queen", "jack", "ten")
hand
## "ace"   "king"  "queen" "jack"  "ten"  

typeof(hand)
## "character"
```
##5.2 Attributes- an attribute is a piece of information you can attach to an atomic vector or any R object. Does not affect values in object, and does not appear when object is displayed. Only some functions use attributes. Ask about attributes with attributes()
```{r}
attributes(die)
```
##Can give attributes names. Look up with names(object) and assign with names(object)<-c(). Manipulating the object will change the vector values, but not the name attributes. To change the names, use the names()<- and reassign names or say <-NULL to remove names.
```{r}
names(die) <- c("one", "two", "three", "four", "five", "six")
names(die)
## "one"   "two"   "three" "four"  "five"  "six" 

attributes(die)
## $names
## [1] "one"   "two"   "three" "four"  "five"  "six"
die
die + 1
##  one   two three  four  five   six 
##    2     3     4     5     6     7
names(die) <- c("uno", "dos", "tres", "quatro", "cinco", "seis")
die
names(die) <- NULL
die
```
##Dimensions. Transform an atomic vector into an n-dimensional array with dim(). For example, make a 2 x 3 matrix out of a vector with dim(object) <-c(rows, columns). Always fills columns first.
```{r}
dim(die) <- c(2, 3)
die
##      [,1] [,2] [,3]
## [1,]    1    3    5
## [2,]    2    4    6
dim(die) <- c(3, 2)
die
##      [,1] [,2]
## [1,]    1    4
## [2,]    2    5
## [3,]    3    6
##Or a 1 x 2 x 3 hypercube (1 row, 2 columns, 3 slices)
dim(die) <- c(1, 2, 3)
die
## , , 1
## 
##      [,1] [,2]
## [1,]    1    2
## 
## , , 2
## 
##      [,1] [,2]
## [1,]    3    4
## 
## , , 3
## 
##      [,1] [,2]
## [1,]    5    6
```
##5.2 Matrices
##Matrices store values as a 2D array. Use matrix(object, nrow= OR ncol=)
```{r}
m <- matrix(die, nrow = 2)
m
```
##Adding byrow = TRUE will fill the matrix by rows
```{r}
m <- matrix(die, nrow = 2, byrow = TRUE)
m
```
##5.4 Arrays
##array() does similar thing as dim(). Do array(atomic vector, vector of dimensions dim=)
```{r}
ar <- array(c(11:14, 21:24, 31:34), dim = c(2, 2, 3))
ar
```
##5.2 exercise: Make a matrix of a royal flush with spades
```{r}
royal<-c("ace", "king", "queen", "jack", "ten", "spades", "spades", "spades", "spades", "spades")
royalm<-matrix(royal, ncol=2)
royalm
```
##5.5 Class
##Changing object dimensions does not change type of object but does change class. Ask for class with class()
```{r}
dim(die) <- c(2, 3)
typeof(die)
##  "double"
 
class(die)
##  "matrix"
```
##Dates and Times: time data type is double, class is "POSIXct" "POSIXt. Get the current time with sys.time()
```{r}
now<-Sys.time()
now

typeof(now)
class(now)
```
##POSIXct saves time as the number of seconds after 12AM Jan 1 1970. Time is a double vector, with the above output and also the seconds after 1970. See this vector using unclass()
```{r}
unclass(now)
```
##Gives you the time at which one million seconds have passed after jan 1 1970
```{r}
mil <- 1000000
mil
## 1e+06
 
class(mil) <- c("POSIXct", "POSIXt")
mil
## "1970-01-12 13:46:40 UTC"
```
##5.5.2 Factors
##Factors are categories with order. Create it with factor(vector). This will recode the data as an integer vector and add a levels attribute to the integer. The vector now has the class called factor. 
```{r}
gender <- factor(c("male", "female", "female", "male"))

typeof(gender)
## "integer"

attributes(gender)
## $levels
## [1] "female" "male"  
## 
## $class
## [1] "factor"

unclass(gender)
## [1] 2 1 1 2
## attr(,"levels")
## [1] "female" "male"  
```
##R uses levels when it displays the factor. They look like character strings but behave as integers. In above example, each 1 is female, each 2 is male, and can go on with more levels.
##Convert a factor to a character string with as.character(factor)
```{r}
as.character(gender)
```
##Exercise 5.4 Write a card
```{r}
AH1<-c("ace", "heart", "1")
AH1
```
##Only one type of data can go into a vector, otherwise it will be coerced into the same type. 
##5.6 Coercion
##If any character string is in a vector, the whole vector becomes character
##If only logicals or numbers, R converts logicals to numbers (True=1, False=0). Same rules apply if you do math with logicals. Can ask R to convert data between types.
```{r}
sum(c(TRUE, TRUE, FALSE, FALSE))
##becomes:
sum(c(1, 1, 0, 0))
as.character(1)
## "1"

as.logical(1)
## TRUE

as.numeric(FALSE)
## 0
```
##5.7 Lists
##Lists group together R Objects. Its elements can be different lengths. list() In results, [[]] is the element number, [] is which subelement
```{r}
list1 <- list(100:130, "R", list(TRUE, FALSE))
list1
```
##Exercise 5.5 Use a list to make a card
```{r}
Card<-list("Queen", "Heart", 7)
Card
```
##5.8 Data Frames
##Data frames group vectors into 2D tables. Each vector is a column. Each column must be the same type of data, but each vector can be different. Make with data.frame(name=vector, name=vector). Vectors must be the same length. The result is a list with the class data frame.
```{r}
df <- data.frame(face = c("ace", "two", "six"),  
  suit = c("clubs", "clubs", "clubs"), value = c(1, 2, 3))
df
```
##Avoid having R turn into a factor by adding stringsAsFactors = FALSE
```{r}
df <- data.frame(face = c("ace", "two", "six"),  
  suit = c("clubs", "clubs", "clubs"), value = c(1, 2, 3),
  stringsAsFactors = FALSE)
```
```{r}
write.csv(deck, file = "cards.csv", row.names = FALSE)
```
#6: R Notation
##Select the specific values from the dataframe you want. Do this with dataframename[index one, index two]
##Indexes can be positive or negative integers, zero, blank spaces, logical values, and names.
###name[row number, column number] to get a specific value
###get the first row with deck[1, c(1,2,3)] or deck[1, 1:3]. Indexing begins with 1!! The first column is 1. 
```{r}
head(deck)
deck[1, 1]
deck[1, c(1, 2, 3)]
```
###Get every index EXCEPT some by putting a negative sign with the ones you want to exclude
```{r}
deck[-(2:52), 1:3]
```
###Blank spaces in indexes will extract every value in that dimension
```{r}
deck[1, ]
```
###Logicals in indexes are just TRUE- return this value or FALSE- don't return this value. 
```{r}
deck[1, c(TRUE, TRUE, FALSE)]
```
###Ask for specific names of rows or columns
```{r}
deck[1, c("face", "suit", "value")]
# the entire value column
deck[ , "value"]
```
##Exercise 6.1 Deal a Card
```{r}
deal <- function(cards) {
  cards[1, ]
}
deal(deck)
```
##6.3 Shuffle the Deck
```{r}
deck2 <- deck[1:52, ]
head(deck2)
```
##Ask for row 2, then row 1, then the rest of the cards
```{r}
deck3 <- deck[c(2, 1, 3:52), ]

head(deck3)
```
##Ask for the rows in a random order using the sample function. Start by getting a random list of numbers, 1:52. Then create a deck which asks for random rows.
```{r}
random <- sample(1:52, size = 52)
random
deck4 <- deck[random, ]
head(deck4)
```
##Exercise 6.2 Shuffle a Deck
##creating the object named random, which is a sample from the entire deck of the entire deck
##selecting 
```{r}
shuffle<- function(cards) {
  random<- sample (1:52, size = 52)
  cards [random, ]
}

deal(deck)
deck2<-shuffle(deck)
deal(deck2)
```
##largerdataset$specificvariable
```{r}
mean(deck$value)
## 7

median(deck$value)
## 7
lst <- list(numbers = c(1, 2), logical = TRUE, strings = c("a", "b", "c"))
lst$numbers
## [1] 1 2

lst$logical
## [1] TRUE

lst$strings
## [1] "a" "b" "c"
lst[1]$numbers
## [1] 1 2
```
##Can't do certain things with a list like sum it, etc. So it is important to be able to ID a vector from a list to do math with it
```{r}
sum(lst[1])
lst$numbers
sum(lst$numbers)
